Drug efficacy prediction in cell lines

### 2019-08-19 docs/manuscript1.Rnw
unpenalized covariates added to rpackage, slight reorganization

	\section{Estimation}\label{sec:estimation}
	\subsection{Variational Bayes}\label{sec:variationalbayes}
	The posterior corresponding to the model described in (\ref{eq:linearmodel})
	and (\ref{eq:priormodel}) is not available in closed form. We therefore 
	approximate the posterior by variational Bayes, where we force the approximate
	posterior 
	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_{1d}^2, \dots, 
	\gamma_{pd}^2) \approx 
	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot \prod_{j=1}^p 
	q(\gamma_{jd}^2)$. For
	notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
	different densities for different inputs.
	Under such a factorisation, the marginal variational posteriors that 
	minimise the Kullback-Leibler divergence of the true posterior to the 
	variational Bayes approximation \cite[]{neal_view_1998} are given by:
	\begin{align*}
    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
    (\bm{\mu}_d, \bm{\Sigma}_d), \\
    q(\gamma_{jd}^2) & \overset{D}{=} \mathcal{GIG}
    \(-1, \lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2, \delta_{jd}\), \\
    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
	$\zeta_d$ contain cyclic dependencies and are iteratively 
	updated by:
  \begin{align*}
    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
    \X \tr \y_d, \\
    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
    (\bSigma_d)_{jj}\] + \lambda_{d}, \\ 
    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
    \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
    \bm{\mu}_d^{(h+1)}\bigg],
  \end{align*}
  until convergence. Here, we set
  \begin{align}
    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), 
    \nonumber\\
    b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
    \sqrt{\frac{\delta_{jd}^{(h)}}
    {\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}}
    \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])}
    {K_1(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])} + 
    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel}
  \end{align}
  where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
  A method for fast and numerically stable calculation of ratios of
  modified Bessel functions of the second kind, as in 
  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
  \ref{sm-sec:ratiosmodifiedbessels}.

@@

  In addition to ADVI, \texttt{rstan} allows the user
  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
  estimates of the model parameters. We compare these MAP estimates together
  with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
  \cite[]{friedman_regularization_2010}, and the ADVI posterior means,
  to the MAP estimates obtained from our model through the correlation with
  and the mean squared distance (MSD) from the MCMC posterior means.

  The interest of this simulation lies in the variational Bayes approximation, 
  so we restrict ourselves to the univariate setting again (i.e., $D=1$). The
  simulation setup is the same as in the dense setting 
  in Section \ref{sec:inversegammamodel}. Similarly, we use the same 
  hyperparameters settings (Table \ref{tab:hyper_set1}. For \texttt{glmnet}
  we either cross-validate the penalty parameter $\lambda$ or match it to the
  true prior variance. We
  repeated the simulations 100 times and report the results in Figure 
  \ref{fig:boxplots_igaussian_res4.2_post1}.

<<boxplots_igaussian_res4.2_post1, fig.cap="(a) \\& (d) Cramér-von Mises criterion, (b) \\& (e) correlation, and (c) \\& (f) MSD between the $\\beta_j$ point estimates of the MCMC and variational bayes posteriors.", out.width="80%", fig.asp=2/3>>=
@

>>>>

  Several extensions of the model are described in [! ref to second paper here].

	\section{Estimation}\label{sec:estimation}
	\subsection{Variational Bayes}\label{sec:variationalbayes}
	The posterior corresponding to the model described in (\ref{eq:linearmodel})
	and (\ref{eq:priormodel}) is not available in closed form. We therefore 
	approximate the posterior by variational Bayes, where we force the approximate
	posterior 
	density to factorise as: $p(\bbeta_d, \sigma_d^2, \gamma_{1d}^2, \dots, 
	\gamma_{pd}^2) \approx 
	Q_d(\cdot) = q(\bbeta_d) \cdot q(\sigma_d^2) \cdot \prod_{j=1}^p 
	q(\gamma_{jd}^2)$. For
	notational convenience, we slightly abuse notation and let $q(\cdot)$ denote
	different densities for different inputs.
	Under such a factorisation, the marginal variational posteriors that 
	minimise the Kullback-Leibler divergence of the true posterior to the 
	variational Bayes approximation \cite[]{neal_view_1998} are given by:
	\begin{align*}
    q(\bm{\beta}_d) & \overset{D}{=} \mathcal{N}_p 
    (\bm{\mu}_d, \bm{\Sigma}_d), \\
    q(\gamma_{jd}^2) & \overset{D}{=} \mathcal{GIG}
    \(-1, \lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2, \delta_{jd}\), \\
    q(\sigma^2_d) & \overset{D}{=} \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
	See SM Section \ref{sm-sec:vbderivations} for the derivations. The variational
	parameters $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, and
	$\zeta_d$ contain cyclic dependencies and are iteratively 
	updated by:
  \begin{align*}
    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
    \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1}, \\
    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + \diag(b_{jd}^{(h)})\]^{-1} 
    \X \tr \y_d, \\
    \delta_{jd}^{(h+1)} & = a_d^{(h)} \[(\bmu^{(h+1)}_{jd})^2 + 
    (\bSigma_d)_{jj}\] + \lambda_{d}, \\ 
    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
    \trace \[ \diag (b_{jd}^{(h+1)}) \bm{\Sigma}_d^{(h+1)}\] + 
    (\bm{\mu}_d^{(h+1)}) \tr \diag (b_{jd}^{(h+1)}) 
    \bm{\mu}_d^{(h+1)}\bigg],
  \end{align*}
  until convergence. Here, we set
  \begin{align}
    a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(n + p + 1)/(2 \zeta_d^{(h)}), 
    \nonumber\\
    b_{jd}^{(h)} & = \E_{Q^{(h)}}(\gamma_{jd}^{-2}) = 
    \sqrt{\frac{\lambda_{d}[(\mathbf{c}_{d}^j) \tr \bm{\alpha}]^2}
    {\delta_{jd}^{(h)}}}
    \frac{K_0(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])}
    {K_1(\sqrt{\delta_{jd}^{(h)}\lambda_{d}}
    [(\mathbf{c}_{d}^j) \tr \bm{\alpha}])} + 
    \frac{2}{\delta_{jd}^{(h)}}, \label{eq:ratiomodifiedbessel}
  \end{align}
  where $K_{\nu}(x)$ denotes the modified Bessel function of the second kind.
  A method for fast and numerically stable calculation of ratios of
  modified Bessel functions of the second kind, as in 
  (\ref{eq:ratiomodifiedbessel}), is given in SM Section 
  \ref{sm-sec:ratiosmodifiedbessels}.

@@

  In addition to ADVI, \texttt{rstan} allows the user
  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
  estimates of the model parameters. We compare these MAP estimates, and the 
  ADVI posterior means,
  to the MAP estimates obtained from our model through the correlation with
  and the mean squared distance (MSD) from the MCMC posterior means.

  The interest of this simulation lies in the variational Bayes approximation, 
  so we restrict ourselves to the univariate setting again (i.e., $D=1$). The
  simulation setup is the same as in the dense setting 
  in Section \ref{sec:inversegammamodel}. Similarly, we use the same 
  hyperparameters settings (Table \ref{tab:hyper_set1}. For \texttt{glmnet}
  we either cross-validate the penalty parameter $\lambda$ or match it to the
  true prior variance. We
  repeated the simulations 100 times and report the results in Figures 
  \ref{fig:boxplots_igaussian_res4.2_post1}-
  \ref{fig:boxplots_igaussian_res4.2_post2}.

  In addition to the performance measures, we have plotted posterior samples 
  histograms of 8 randomly selected model parameters in setting 1, 
  along with the esimated
  marginal variational Bayes posteriors, and the MAP estimates in Figure 
  \ref{fig:hist_igaussian_fit4}.

<<boxplots_igaussian_res4.2_post1, fig.cap="Comparison of different approximations in (a)-(c) simulation setting 1, (d)-(f) setting 2, and (g)-(i) setting 3, through Cramér-von Mises criterion between the approximate and MCMC posteriors, correlation and MSD between the approximate and MCMC posterior means.", out.width="80%", fig.asp=1>>=
@

<<boxplots_igaussian_res4.2_post2, fig.cap="Comparison of different approximations in (a)-(c) simulation setting 4, (d)-(f) setting 5, and (g)-(i) setting 6, through Cramér-von Mises criterion between the approximate and MCMC posteriors, and correlation and MSD between the approximate and MCMC posterior means.", out.width="80%", fig.asp=1>>=
@

<<hist_igaussian_fit4, fig.cap="Histogram of posterior MCMC samples of 8 randomly selected model parameters in setting 1, along with the estimated variational posteriors and MAP estimates.", out.width="80%", fig.asp=1>>=
@

### 2019-08-19 docs/manuscript2.Rnw
unpenalized covariates added to rpackage, slight reorganization

	\section{Methods}
	\subsection{Simultaneous equations model}\label{sec:SEM}
	Assume we have continuous efficacy measures $y_{id}$ for cell lines 
	$i=1,\dots, n$, centred per drug $d$. We predict efficacy from molecular 
	features 
	$x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & 
	\dots & x_{ip} \end{bmatrix} \tr$. We have an \textit{a priori} (possibly 
	informative) partitioning 
	of the omics features into $g=1, \dots, G$ non-overlapping groups.
	For convenience, we let 
	$g(j)$ denote the group index of molecular feature $j$. Furthermore, we have 
	characteristics of the drugs available in the form of external variables 
	$\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D 
	\end{bmatrix} \tr$.

	We regress the drug efficacies on the molecular feature effects $\beta_{jd}$:
	\begin{subequations}
		\begin{align}
		  y_{id} & = \x_i \tr \bbeta_d + \epsilon_{id}, \label{eq:linearmodel}\\
		  \text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
		\end{align}
	\end{subequations}
	where $p$-dimensional $\bbeta_d$ are the drug-specific omics feature effects. 
	Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ 
	linear simultaneous equations.

	\subsection{Bayesian prior model}
	We capture the uncertainty in the parameters through a Bayesian prior model
	and endow the parameters with the following prior:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} & \sim \mathcal{N}_p (0, \phi^2_{g(j)} \gamma_{jd}^2 
		  \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}( 
		  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}, \lambda_d), \\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\theta, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\theta$ and shape $\lambda$. The different molecular feature types
	are given their own weights through the $\phi^2_{g}$, while the external 
	outcome-feature data comes in through 
	our mean model for the $\gamma_{jd}^2$ hyperprior model: 
	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. Both the $\phi^2_{g}$ and
	$\bm{\alpha}$ are estimated in a data-driven manner. More details on the model
	itself, as well as its estimation, may be found in 
	the SM and in [! ref to first paper here].

>>>>

	\section{Methods}
	\subsection{Simultaneous equations model}\label{sec:SEM}
	Assume we have continuous efficacy measures $y_{id}$ for cell lines 
	$i=1,\dots, n$, centred per drug $d$. We predict efficacy from molecular 
	features 
	$x_{ij}$, $j=1,\dots, p$, collected in $\x_i = \begin{bmatrix} x_{i1} & 
	\dots & x_{ip} \end{bmatrix} \tr$. Furthermore, we have 
	characteristics of the drugs available in the form of external variables 
	$\mathbf{C} = \begin{bmatrix} \mathbf{c}_1 & \dots & \mathbf{c}_D 
	\end{bmatrix} \tr$.

	We regress the drug efficacies on the molecular feature effects $\beta_{jd}$:
	\begin{subequations}\label{eq:likelihood}
		\begin{align}
		  y_{id} & = \x_i \tr \bbeta_d + \epsilon_{id}, \label{eq:linearmodel}\\
		  \text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
		\end{align}
	\end{subequations}
	where $p$-dimensional $\bbeta_d$ are the drug-specific omics feature effects. 
	Note that (\ref{eq:linearmodel}) gives rise to a system of $d=1, \dots, D$ 
	linear simultaneous equations.

	\subsection{Bayesian prior model}
	We capture the uncertainty in the parameters through a Bayesian prior model
	and endow the parameters with the following prior:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \sigma^2_d & \sim \mathcal{N}_p 
		  (0, \gamma_{jd}^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}( 
		  [(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}, \lambda_d), \\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\theta, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\theta$ and shape $\lambda$. The external 
	outcome-feature data comes in through 
	our mean model for the $\gamma_{jd}^2$ hyperprior model: 
	$[(\mathbf{c}_d^j) \tr \bm{\alpha}]^{-1}$. The
	$\bm{\alpha}$ and $\lambda_d$ are estimated in a data-driven manner. 
	More details on the model
	itself, as well as its estimation, may be found in 
	the SM and in [! ref to first paper here].

	\subsection{Tissue effect}
	The cell lines used in drug response models are often taken from different
	tissues. In addition, other clinical covariates are often available. 
	For unbiased omics feature effects, one may wish to account for these, 
	without shrinkage of their effects. We do that by adding such unpenalized 
	covariates 
	$\mathbf{z}_{i} = \begin{bmatrix} z_{i1} & \cdots & z_{iT} \end{bmatrix} \tr$ 
	to model (\ref{eq:likelihood}):
  \begin{align*}
    y_{id} & = \x_i \tr \bbeta_d + \mathbf{z}_i \tr \bm{\psi} + 
    \epsilon_{id}, \\
    \text{with } & \epsilon_{id} \sim \mathcal{N}(0, \sigma_d^2),
	\end{align*}
	with $\bm{\psi} = \begin{bmatrix} \psi_1 & \cdots & \psi_T \end{bmatrix} \tr$ 
	the unpenalized covariates effects. We assign $\bm{\psi}$ a flat prior, which 
	alters the estimation as described in SM Section [! ref to section here].
